Réponse et résilience de la biodiversité des forêts tropicales après perturbation

### 2018-09-05 docs/Manuscript.tex
texte complet

\usepackage[style=authoryear-ibid,backend=bibtex,citestyle=verbose-inote,isbn=false,backref=true,giveninits=true,uniquename=init,maxcitenames=2,maxbibnames=150,sorting=nyt,sortcites=false]{biblatex}
\addbibresource{References.bib}

@@

@@

Article 3: Analyse du recrutement, support de la trajectoire des
communautés

>>>>

\usepackage[style=authoryear-ibid,backend=bibtex,citestyle=verbose-inote,isbn=false,backref=true,giveninits=true,uniquename=init,maxcitenames=2,maxbibnames=150,sorting=nyt,sortcites=false]{biblatex}
\addbibresource{References.bib}
\addbibresource{packages.bib}

@@

\subsection{Abstract}\label{abstract}

Assess the fate of Neotropical forests requires to accurately measures
forest diversity and reliably monitor forest communities. The costs of
botanical inventories and the taxonomic complexity of Neotropical
forests make forest inventories in vernacular names the most efficient
approach today, although these hold high botanical uncertainty and limit
the accuracy of diversity measures. Several methods were proposed to
compensate these botanical uncertainties but none reliably assessed
functional and fine-scale diversity surveys. We developed a polyvalent
diversity estimator workable in numerous specific cases based on the
propagation of botanical uncertainties. The estimator was calibrated
with a large neotropical inventory and the simulations of uncertainty
and sampling effort gradients allowed to determined an ideal inventory
protocol optimizing the costs and the accuracy of forest inventories.\\
Our study first advocated of necessity of real inventories and the
inescapable recourse to taxonomists to ensure reliable diversity
estimations. An ideal inventory protocol based on a sampling effort of 3
000 trees and on an identification effort of 80\% of the species was
identified and ensured diversity estimations with a 10\% error.

\subsection{Introduction}\label{introduction}

The variety of tree species, their assemblages in space and their
dynamics in time are determinant of forests productivity and functioning
\autocite{Cardinale2012}. Preserve tree diversity is crucial to maintain
forests functioning and services, specifically in hyper-diverse tropical
forests where the biodiversity is as threatened as it is valuable and
unexplored \autocite{Barlow2018}. Handling the conservation and
management of tree diversity requires setting sensible protection areas
and sustainable forest management calibrated according to diversity
patterns in space and time and their determinants
\autocites{Margules2000}{Purvis2000}{Gibson2011a}{FAO2014}{Sist2015}.

Correctly measure, map and manage forests biodiversity require accurate
and large forest monitoring. The precision of forest inventories,
though, is often limited by their significant cost in terms of time,
money, and logistic \autocites{Feeley2011}{Baraloto2013}. Sampling
methods were optimized to minimize these costs and maximize inventory
accuracy.Some approaches would restrict inventories to some DBH or
height classes, to specific taxa, or would opt for inventories at family
or genus level. These methods efficiently translated biodiversity
patterns at regional scales and along wide ecological gradients
\autocites{Steege2000}{Higgins2004}{Rejou-Mechain2011}{Pos2014}.
However, these methods were either limited to small areas (under 1ha),
sometimes remained biased or holding significant uncertainty, and
usually proved limited to detect subtle diversity aspects and to
desentangle richness from equitability parameters
\autocites{Phillips2003a}{Baraloto2013}[
]{Guitet2014b}{Vellend2008}{Prance1994}. Another approach proposed to
use inventories in vernacular names instead of botanical species.
Vernacular names indeed are easier to attribute, more common and usually
do not require vouchers collection or posterior botanical
identification. The reliability of vernacular names may be high at genus
level, but this proved highly variable across tropical regions: while
this reliability was estimated around 60-70\% in French Guiana
\autocites{Hawes2012}{Guitet2014b} to ranges from 32\% to 67\% in
Central Africa \autocite{Rejou-Mechain2011}. The multiple and variable
associations between botanical and vernacular names then entail
significant botanical uncertainties that should not be ignored
\autocite{Oldeman1968}. Besides, rough vernacular inventories would not
allow functional and phylogenetic approaches, that require
identification at the botanical species to comply with phylogenetic and
functional database. However the approach through vernacular names
deserves further attention. First, it gives the opportunity to analyze
pre-logging inventories conducted in large areas by logging companies.
Second, as exhaustive inventories, they allow some post-process based on
vernacular/botanical names association and allow the building of
reliable diversity estimators
\autocites{TerSteege2006}{Feldpausch2006}{Rejou-Mechain2008}{Rejou-Mechain2011}.
Following this idea \textcite{Guitet2014b} proposed a framework
propagating vernacular names taxonomic uncertainties in diversity
measures. The propagation framework was based on Monte-Carlo processes
estimating forest diversity from the vernacular-botanical name
association. These association combined prior information from both
general taxa-abundance correspondence table \autocite{Molino2009} and
reference field inventories. The framework successfully rendered the
ranking of plots diversity, but remained restricted to large
environmental gradient and for highly different communities
\autocites{Guitet2014b}{Guitet2013}. In this study we offer to refine
this framework and adapt it to diversity estimation at smaller spatial
scales. The following diversity estimator is based on the specific case
of the studied community and the inventory protocol. The diversity
estimator besides suits all inventories whatever the ratio of botanical
determination, \emph{i.e.} ratio of vernacular compared to botanical
names. It besides suits experimental specific as well as pre-logging
inventories where only the commercial or most recognizable species are
identified at species level.

Such diversity estimator allows maximizing the accuracy of diversity
measures while minimizing the sampling effort, \emph{i.e.} the size of
inventoried communities and the number of accurately identified species.
In this perspective we thought to calibrate an ideal inventory protocol
optimized in terms of sampling effort and determination degree. From a
real inventory, with complete vernacular and botanical identifications,
we simulated ranges of sampling efforts and identification degrees along
which we examined the bias and variability of the diversity estimator.

In this study we \emph{(i)} redesigned a diversity estimator based on a
Bayesian framework accounting for both general taxa-association tables
and specific field inventories, and \emph{(ii)} applied the estimator to
a real Neotropical forest inventory to determine the sampling effort and
determination degree of an ideal inventory protocol.

\subsection{Methods}\label{methods}

\subsubsection{Study community}\label{study-community}

We based our analyses on the inventory of a Neotropical rainforest, from
the Paracou Research Station in French Guiana (5°18'N and 52°53'W). The
experimental site stands in a lowland tropical rainforest with a flora
dominated by \emph{Fabaceae}, \emph{Chrysobalanaceae},
\emph{Lecythidaceae} and \emph{Sapotaceae} families. Mean mean annual
temperature is 26°C. and the mean annual precipitations average
\(2980 mm.y^-1\) (30-y period) with a 3-months dry season
(\(< 100 mm.months-1\)) from mid-August to mid-November and a one-month
dry season in March \autocite{Wagner2011}. Elevation ranges between 5
and 50 m and soils correspond to thin acrisols over a layer of
transformed saprolite with low permeability, generating lateral drainage
during heavy rains \autocite{IUSSWorkingGroupWRB2015}. We used the 2015
inventory of six permanent plots of undisturbed forest (6.25ha each,
37.5ha inventoried in total). During inventories trees are identified
first with a vernacular name assigned by the forest worker team, and
afterward with a scientific name assigned by botanists during regular
botanical campaigns. The community inventoried ancompasses 22 904 trees
belonging to 375 species and 63 families, identified by 290 different
vernacular names. The initial taxonomic uncertainty was 3\% of the
community, \emph{i.e.} the proportion of trees not identified with a
botanical name.

\subsubsection{Diversity measures}\label{diversity-measures}

Among the large panel of diversity indices we examined here the family
of q-generalized (Tsallis) entropy, widely adopted to assess all aspects
of taxonomic, functional and phylogenetic diversities. The Tsallis
diversity indices derive from a general formula, modulated by an order q
emphasizing species frequency \eqref{eq:TsallisEntropy}.

\begin{equation}
^qD = \sum_{i=1}^{N}{\left( p_i^q \right)^{\frac{1}{1-q}} }
\label{eq:TsallisEntropy}
\end{equation}

In the diversity formula, species relative abundance \(p_i\) in a
community of \(N\) species is raised at the power \(q\) that is the
order of the diversity. The higher the order \(q\), the higher the
emphasis on common vs.~rare species, so browsing a range of order \(q\)
corresponds assess a gradient balance between richness and evenness. The
formula retrieves species richness for \(q = 0\) , Shannon diversity for
\(q = 1\) where richness and evenness are equally accounted for and
Simpson diversity, that can be undestood as the diversity of common
species, for \(q = 2\). The Tsallis diversity indices would eventually
be converted into equivalent number of species in our framework. The
conversion in equivalent number of species, through Hill transformation,
allows understandable analysis and comparisons among communities
\autocites{Hill1973}{Keylock2005}{Jost2006}.

\subsubsection{Diversity estimator}\label{diversity-estimator}

The estimation framework is based on the diversity distribution measured
on theoretical, fully determined communities. Theoretical inventories
are simulated 1 000 times from the real incomplete inventory, through
the replacement by a Monte-Carlo scheme of vernacular names by botanical
ones.

The vernacular-botanical replacement are based on the association
probability between each vernacular names and the botanical names
inventoried. For each vernacular name the association model follows a
multinomial distribution
\(M([s_1, s_2, …, s_N] ,[\alpha_1, \alpha_2,…, \alpha_N])\), with
\([\alpha_i]\) the association probability of botanical name \(s_i\)
with the vernacular name.

The association probability vectors \([\alpha_v]\) were determined with
a Bayesian framework based on the combination of botanical expertise and
observed associations. First, the estimation of \([\alpha_v]\) accounted
for prior information from experts' knowledge in the form of a general
taxa-association table listing all botanical names likely corresponding
to the vernacular name \(v\). From this general table, the probability
\(\lambda_i={}^1/m_v\) was attributed to each of the \(m_v\) botanical
names with a confirmed association with \(v\). When no association was
established the probability \(\lambda_i={}^\epsilon\big/_{N-m_v}\) was
attributed to the botanical name, with \(\epsilon\) standing for a
background noise set to 0.01 here. Second, the estimation of
\([\alpha_v]\) accounted for observed inventories giving real
association frequencies \(\phi_i\) between \(v\) and the \(m_v'\)
botanical names with observed association. Similarly, the association
probability \(\lambda_i={}^\epsilon\big/_{N-m_v'}\) was attributed to
botanical names with no observed association. The final \([\alpha_v]\)
distribution was modeled by a Multinomial-Dirichlet scheme combining the
two vectors \([\lambda^v]\) and \([\phi^v]\) \autocite{McCarthy2007}.

To test the relevance of the general table and observed inventories
information, we tested a range of weighting \(w\). Assuming a
distribution of \([\phi^v]\) conditionally to \([\alpha^v]\) the
weighting returned the formula \eqref{eq:weighting}.

\begin{equation}
[\alpha_i^v]: 
\Big[\alpha_i^v | _{(1-w)\lambda_i^v ,w.\phi_i^v}\Big] =Dirichlet\Big((1-w)\phi_i^v+w.\lambda_i^v\Big)
\label{eq:weighting}
\end{equation}

When \(w=0\) only observed inventories were considered, when \(w=0.5\)
both information were equally accounted for and when \(w=1\) only the
general taxa-association table was considered.

\subsubsection{Simulation of determination and sampling effort
gradients}\label{simulation-of-determination-and-sampling-effort-gradients}

The estimator was calibrated in comparing several methods for the
vernacular/botanical association probability (corresponding to different
values of \(w\), the balance between general table and observed
inventories). The performance of the estimator was examined regarding
its bias, \emph{i.e.} the difference between the estimation and the real
diversity, and its variability, \emph{i.e.} 95\% confidence interval
\autocite{Baltanas2009}. For each computation method the performance of
the estimator was examined along a determination gradient (corresponding
to an increasing number of species only identified in vernacular name).
As rare species had more chance to be undetermined (Kendall test,
\(\tau = -0.46, p < 10^-16\)), the trial of ignored determination
followed botanical names abundance (\(p_{undetermined}=f_i^{-0.1}\),
with \(f_i\) botanical name frequency).

Different inventory specific cases were then tested in examining the
bias and variability of the estimator along a sampling effort gradient
(corresponding to an increasing number of trees, from 500 to 22 000
trees, used to compute the vernacular/botanical association
probability). Along the sampling effort gradient the estimations were
performed on the fully undetermined inventory, \emph{i.e.} without any
botanical identification.

\subsection{Results}\label{results}

\subsubsection{The reponse to determination effort, and the design of an
ideal
framework}\label{the-reponse-to-determination-effort-and-the-design-of-an-ideal-framework}

Along the indetermination gradient, when considering both general
taxa-association table and observed inventory the diversity was
increasingly overestimated (Fig. \ref{fig:UncertGrad}). This
overestimation increased with the order of diversity q, while it was not
significant for the richness (\(q=0\)), the overestimation reached 45\%
of the real diversity for Shannon diversity (\(q = 1\)) and it reached
57\% of the real diversity for the Simpson diversity (\(q = 2\)).

When only considering the general taxa-association table the richness
(\(q=0\)) was underestimated (reaching a 50\% underestimation), while
both Shannon and Simpson diversities were overestimated (respectively
reaching underestimations of 67\% and 125\%).

When only considering the observed inventory the estimator remained
slightly biased but it did not exceed 15\% of the real diversity for any
order of diversity.

A bootstrap of the 100 simulations for each specific case and diversity
order showed a stabilization of variances after 60 simulations.

{\centering \includegraphics[width=1\linewidth]{Manuscript_files/figure-latex/UncertGrad-1}

\caption{Richness, Shannon and Simpson estimator bias and 95\%confidence interval along an uncertainty gradient with the association frequencies computed from (a) only expert prior, (b) both expert and observation prior and (c) only observation prior.}\label{fig:UncertGrad}
\end{figure*}

\subsubsection{Calibrating the sampling
effort}\label{calibrating-the-sampling-effort}

Along the sampling effort gradient from 500 to 22 000 trees, the
richness estimator remained negatively biased but the confidence
interval did not exceed 7\%. The Shannon and Simpson were less biased,
for 3 000 trees inventoried the Shannon diversity bias fell to 15\%
while the bias of Simpson estimator fell to 6\% (Fig.
\ref{fig:SEgradient}).

{\centering \includegraphics[width=1\linewidth]{Manuscript_files/figure-latex/SEgradient-1}

\caption{Richness, Shannon and Simpson estimation (upper panels) and bias compared to real diversity (lower panels) along a sampling effort gradient. Shaded areas are the 95\% confidence intervals, vertical plain line stands for the points at 3 000 trees.}\label{fig:SEgradient}
\end{figure*}

The precision and bias of the estimator were eventually tested for the
recomended sampling effort of 3 000 trees. In this case the Shannon and
Simpson biases remained lower than 10\% and the Richness bias was below
10\% until 20\% of undetermined species.

{\centering \includegraphics[width=1\linewidth]{Manuscript_files/figure-latex/UncertGradLim-1}

\caption{Richness, Shannon and Simpson estimators along a taxonomic uncertainty gradient based on field inventories of 3 000 trees,  with shaded areas the 95\% confidence interval.}\label{fig:UncertGradLim}
\end{figure*}

\subsection{Discussion}\label{discussion}

\subsubsection{Inescapable taxonomists}\label{inescapable-taxonomists}

The method developed in the line of \textcite{Guitet2014b} to propagate
the taxonomic uncertainty of vernacular names in diversity mesures
provided a reliable estimator for diversity indices, workable for all
diversity order and adaptable for functoinal diversity. The use the
general taxonomic-association table proved to systematically
overestimate the diversity. In these tables the vernacular/botanical
association probabilities were independent of botanical name abundances,
so rare vernacular names were indifferently replaced by rare or abundant
botanical names. As a result, the abundance of rare species were
inflated at the expense abundant ones, which overestimated the
equitability. Contrastingly the use of observed inventories that account
for the abundance of botanical names proved more reliable. The rescourse
to taxonomists and pre-inventories proved unavoidable to correctly
estimate and therefore manage forest biodiversity.

\subsubsection{Calibration of an optimized inventory
protocol}\label{calibration-of-an-optimized-inventory-protocol}

The performance of the estimator, regarding its bias and variability,
along the determination and sampling efforts gradients highlighted the
difficulty to assess communities richness. Whatever the inventory
protocol the richness indeed remained significantly biased, as already
suggested in previous analysis comparing several inventory methods
\autocite{Higgins2004}.

Conversely, the Shannon and Simpson diversity estimations proved less
biased, thus allowing the estimator to detect small variation of
community equitability. This would be a key to value existing
inventories and ease future protocols as subtle time and spatial
diversity variations often invlove changes in community abundance
distribution rather than richness
\autocites{Baraloto2012a}{Berry2008a}{Cannon1998}{Plumptre1996}.

An optimized protocol maximizing the estimator performance on the one
hand and minimizing the determination and sampling efforts on the other
hand corresponding to the pre-inventory of at least 3 000 trees and the
determination of 80\% of the botanical species.

\subsection{Conclusion}\label{conclusion}

The diversity estimator developed in this paper (i) proved relevant to
measure tropical forest diversity at small time and spatial scale, (ii)
allowed to integrate the specificities of local forests and working
team, and (iii) was adaptable for both taxonomic and functional
diversity at all order. Examining the estimator performance highlighted
the inescapable rescourse to taxonomists and to minimum real
inventories: an initial inventory of 3 000 trees with 80\% of the
species identified allowed an estimation accurate at 10\% of Shannon and
Simpson diversities. The diversity estimator allows integrating and is
thus adaptable to all specific case.

@@

\section*{Post-Disturbance Tree Community Trajectories in a Neotropical
Forest}\label{post-disturbance-tree-community-trajectories-in-a-neotropical-forest}
\addcontentsline{toc}{section}{Post-Disturbance Tree Community
Trajectories in a Neotropical Forest}

\subsection{Abstract}\label{abstract-1}

Understanding the ecological rules underlying the maintenance of
tropical forests biodiversity, structure, functioning and dynamics is
urgent to anticipate their fate in the global change context. The huge
diversity of tropical forests is often assumed to be regularly reshaped
by natural disturbance yielding a diversity peak at intermediate
intensity. This intermediate disturbance hypothesis (IDH), though,
remains debated and the controversy questions the extent of communities
resilience regarding their taxonomic and functional facets. To
disentangle the ecological processes driving community response to
disturbance, we analyzed the tree community trajectories over 30 years
following a disturbance gradient in a Neotropical forest. Specifically,
we examined community functional and taxonomic trajectories with regards
to diversity, composition and redundancy. Functional trajectories were
drawn based on 7 leaf, stem and life-history traits. We highlighted the
cyclic recovery of community taxonomic and functional composition. While
pre-disturbance taxonomic differences were maintained over time,
functional composition trajectories were quite similar among
communities. The IDH did predict communities taxonomic diversity
response while functional diversity was enhanced whatever the
disturbance intensity. Although consistent, the recovery of community
composition, diversity and redundancy remained unachieved after 30
years. This acknowledged the need of decades-long cycles with no
disturbance to ensure a complete recovery, and questioned tropical
forest community resilience after repeated disturbances.

\textbf{Keywords}: Community Ecology, Determinants of Plant Community
Diversity and Structure, Disturbance Trajectories, Intermediate
Disturbance Hypothesis, Long-term Resilience, Neotropical Forests,
Taxonomic and Functional Biodiversity

\subsection{Introduction}\label{introduction-1}

The large areas covered with tropical forests worldwide hold crucial
environmental, economic and social values. They provide wood and
multiple non-timber forest products, shelter a diversified fauna,
regulate the local and regional climates, the carbon, water and nutrient
cycles, and ensure cultural and human well-being. The growing demand in
forests products together with current global changes increases the
pressure on remaining natural forests \autocite{Morales-Hidalgo2015} and
threatens the maintenance and dynamics in space and time of communities
structure, composition and functioning \autocite{Anderson-Teixeira2013}.

In tropical forests, ecological communities are regularly re-shaped by
natural disturbance events changing both the abiotic environment,
through the fluxes of light, heat and water
\autocite{Goulamoussene2017}, and the biotic interactions such as
competition among species \autocite{Chesson2000}. One of the cornerstone
of tropical forest ecology is to understand the processes and drivers of
ecosystems response to disturbance \autocite{Chazdon2003a}. For now,
this has been largely studied through forest structural parameters such
as aboveground biomass, tree height or stem density
\autocites{Piponiot2016}{Rutishauser2016} that are rapid and convenient
to measure. These structural parameters have been successfully modeled,
giving important insights into the recovery of ecosystem processes and
services \autocite{Herault2018}. However the response of forests
diversity and composition remains unclear, albeit it determines the
productivity, stability and functioning of ecosystems
\autocites{Tilman2014}{Liang2016}. In the short-term, moderate
disturbance may lead to positive impacts on communities diversity, an
idea formalized by the intermediate disturbance hypothesis (IDH) stating
a maximized species diversity when disturbance intensity is not too high
\autocites{Molino2001}{Kariuki2006a}.

Validations of the IDH though remain scarce in the long-term and mainly
rely on the analysis of taxonomic richness \autocite{Molino2001}.
Taxonomic richness alone, however, gives limited or misleading
information on forests recovery and functioning
\autocite{Chaudhary2016}. More ecological-meaningful analysis would
couple richness with (i) evenness, that would reveal the changes in the
species abundance distribution and thus the underlying ecological
processes, and (ii) composition that is crucial for conservation issues
\autocites{Lavorel2002}{Bellwood2006}. Furthermore, a functional
approach accounting for species biological attributes would directly
link communities diversity, composition and redundancy to ecosystem
functioning and to its environmental constraints
\autocites{Violle2007b}{Baraloto2012a}. In that respect, the functional
trait-based approach that focus on major traits related to species
ecology and mediate species performance in a given environment was
successfully adopted \autocite{Diaz2005}. For instance, the functional
approach revealed in tropical rainforests the deterministic processes
entailing, after disturbance, a functional shift from a dominance of
``conservative'' slow-growing species dealing with scarce resources to
``acquisitive'' fast-growing species with rapid and efficient use of
abundant resources \autocites{Reich2014}{Herault2011}. This shift is
translated into the trajectories of key functional traits related to
resource acquisition (leaf and stem traits) and life-history traits
(seed mass, maximum size)
\autocites{Wright2004}{TerSteege2006}{Westoby2006a}{Chave2009b}.
Eventually a complete overview of community response to disturbance
would encompass the changes in functional redundancy, that quantifies
the amount of shared trait values among species \autocite{Carmona2016}.
The high functional redundancy of hyper-diverse tropical forests
\autocite{Bellwood2006} mitigates the impacts of species removal on
ecosystem functioning and determines the resilience of communities after
disturbance \autocites{Elmqvist2003}{Diaz2005}.

In this study, we monitored over 30 years the response of 75 ha of
Neotropical forest plots set up on a gradient of disturbance intensity,
from 10 to 60\% of ecosystem above-ground biomass (AGB) loss. We made
use of a large functional traits database encompassing major leaf, stem
and life-history traits in order to draw the taxonomic and functional
trajectories in terms of richness, evenness, composition and redundancy.
Specifically, we (i) elucidated community taxonomic and functional
recovery and the underlying ecological processes, (ii) clarified the
validity of the IDH in the long term for tropical forest and its
translation into different trajectories in time, and (iii) questioned
community recovery time.

\subsection{Material and Methods}\label{material-and-methods}

\subsubsection{Study site}\label{study-site}

Paracou station in French Guiana (5\textdegree 18'N and
52\textdegree 53'W) is located in a lowland tropical rain forest in a
tropical wet climate with mean annual temperature of 26\textdegree C,
mean annual precipitation averaging 2980 mm.y\textsuperscript{-1} (30-y
period) and a 3-month dry season (\textless{} 100
mm.month\textsuperscript{-1}) from mid-August to mid-November, and a
one-month dry season in March \autocite{Wagner2011}. Elevation ranges
between 5 and 50 m and soils correspond to thin acrisols over a layer of
transformed saprolite with low permeability generating lateral drainage
during heavy rains.

The experiment is a network of twelve 6.25ha plots that underwent a
disturbance gradient of three logging, thinning and fuelwood cutting
treatments (Table \ref{tab:Tab1}) according to a randomized plot design
with three replicate blocks of four plots. The disturbance corresponds
to averages of 10 trees removed per hectare with a diameter at 1.3 m
height (DBH) above 50 cm for treatment 1 (T1), 32 trees/ha above 40 cm
DBH for treatment 2 (T2) and 40 trees/ha above 40 cm DBH for treatment 3
(T3). Treatments T2 and T3 besides included the thinning of trees by
poison girdling \autocite{Schmitt1990}. The disturbance intensity was
measured as the percentage of aboveground biomass (\%AGB) lost between
the first inventory in 1984 and five years after disturbance
\autocite{Piponiot2016} estimated with the BIOMASS R package
\autocite{Biomass2018}.

\begin{longtable}[t]{l|l|l|l|l}
\caption{\label{tab:Tab1}Intervention table, summary of the disturbance intensity for the 4 plot treatments in Paracou.}\\
\hline
Treatment & Timber & Thinning & Fuelwood & \%AGB lost\\
\hline
Control & - & - & - & 0\\
\hline
T1 & DBH $\geq$ 50 cm, commercial species, $\approx$ 10   $trees.ha^{-1}$ & - & - & $[12-33]$\\
\hline
T2 & DBH $\geq$ 50 cm, commercial species, $\approx$ 10  $trees.ha^{-1}$ & DBH $\geq$ 40 cm, non-valuable species, $\approx$ 30   $trees.ha^{-1}$ & - & $[33-56]$\\
\hline
T3 & DBH $\geq$ 50 cm, commercial species, $\approx$ 10  $trees.ha^{-1}$ & DBH $\geq$ 50 cm, non-valuable species, $\approx$ 15  $trees.ha^{-1}$ & 40 cm $\leq$ DBH $\leq$ 50 cm, non-valuable species,\ $\approx$ 15 $trees.ha^{-1}$ & $[35-56]$\\
\hline
\end{longtable}

\subsubsection{Inventories protocol and dataset
collection}\label{inventories-protocol-and-dataset-collection}

The study site corresponds to a tropical rainforest typical of the
Guiana Shield with a dominance of Fabaceae, Chrysobalanaceae,
Lecythidaceae and Sapotaceae. In the twelve experimental plots, all
trees above 10 cm DBH have been mapped and measured annually since 1984.
Trees are first identified with a vernacular name assigned by the forest
worker team, and afterward with a scientific name assigned by botanists
during regular botanical campaigns. In 1984, specific vernacular names
were given to 62 commercial or common species whereas more infrequent
ones were identified under general identifiers only distinguishing trees
and palm trees. From 2003, botanical campaigns have been conducted every
5 to 6 years to identify all trees at the species level but
identification levels still varied among plots and campaigns.

This variability of protocols in time raised methodological issues as
vernacular names usually correspond to different botanical species. It
resulted in significant taxonomic uncertainties that had to be
propagated to composition and diversity metrics. The uncertainty
propagation was done through a Bayesian framework reconstituting
complete inventories at genus level from real incomplete ones on the
basis of vernacular/botanical names association. Vernacular names were
replaced through multinomial trials based on the association probability
\(\big[\alpha_1, \alpha_2,..., \alpha_V\big]\) observed across all
inventories between each vernacular name \emph{v} and all species
\(\big[s_1, s_2,..., s_N\big]\):

\begin{align}
M_v\Big(\big[s_1, s_2,..., s_N\big],\big[\alpha_1, \alpha_2,..., \alpha_V\big]\Big) \nonumber
\end{align}

See Supplementary Materials -Fig. S1 and \textcite{Aubry-Kientz2013} for
the detailed methodology.

Six functional traits representing leaf economics (leaves thickness,
toughness, total chlorophyll content and specific leaf area) and stem
economics (wood specific gravity and bark thickness), and life-history
traits (maximum specific height and seed mass) came from the BRIDGE
project \footnote{\url{http://www.ecofog.gf/Bridge/}}. Trait values were
assessed from a selection of individuals located in nine permanent plots
in French Guiana, including two in Paracou, and comprised 294 species
pertaining to 157 genera. Missing trait values (10\%) were filled using
multivariate imputation by chained equation \autocite{Mice2011}.
Imputations were restricted within genus or family when samples were too
scarce, in order to account for the phylogenetic signal. Whenever a
species was not in the dataset, it was attributed a set of trait values
randomly sampled among species of the next higher taxonomic level (same
genus or family). As seed mass information was classified into classes,
no data filling process was applied and analyses were restricted to the
414 botanical species recorded.

All composition and diversity metrics were obtained after 50 iterations
of the uncertainty propagation framework.

\subsubsection{Composition and diversity
metrics}\label{composition-and-diversity-metrics}

To counter taxonomic uncertainties due to the variability of botanical
identification levels (in space) and protocols (in time), the taxonomic
composition and diversity analysis were conducted at the genus level.
Taxonomic and functional trajectories of community composition were
followed in a two-dimensional NMDS ordination space. Two NMDS using
abundance-based (Bray-Curtis) dissimilarity measures were conducted to
map either taxonomic or functional composition, the later based on the 7
leaf, stem and life history traits (without seed mass classes).
Trajectories along time were reported through the Euclidean distance
between the target inventories and the reference inventories in 1989,
\emph{i.e.} 2 years after disturbance. Univariate trajectories of the
leaf, stem and life-history traits were also visualized with the
community weighted means (CWM) \autocite{Diaz2007}. Species seed mass
corresponded to 5 classes of increasing mass, seed mass trajectories
were therefore reported as the proportion of each class in the
inventories (Supplementary materials).

The taxonomic diversity was reported through species richness and
evenness, \emph{i.e} the Hill number translation of the Simpson index
\autocite{Hill1973}. These indices belong to the set of HCDT or
generalized entropy, respectively corresponding to the 0 and 2 order of
diversity (q), recommended for diversity studies \autocite{Marcon2015b}.
The functional diversity was reported using the functional richness and
functional evenness, \emph{i.e} Rao index of quadratic entropy which
combines species abundance distribution and average pairwise
dissimilarity based on species functional traits.

The impacts of initial disturbance were tested with the spearman rank
correlation between the extrema of taxonomic and functional metrics
reached over the 30 years and the initial \%AGB loss. They were besides
analyzed through polynomial regression between (i) taxonomic and
functional richness and evenness and (ii) the initial \%AGB loss at 10,
20 and 30 years after disturbance.

The functional redundancy was measured as the overlap among species in
community functional space \autocite{Carmona2016}. The samples of the
trait database were first mapped in a 2-dimensional plan with a PCA
analysis. Then, multivariate kernel density estimator associated with
individual trees returned species traits probability distribution (TDP).
Species TDP weighted by species abundance were eventually summed for
each community. Community functional redundancy was the sum of TDPs
overlap, expressed as the average number of species that could be
removed from without reducing the functional space (Supp. Mat. - Fig. S1
for a more comprehensive scheme).

\subsection{Results}\label{results-1}

\subsubsection{Communities Composition}\label{communities-composition}

From 1989 (2 years after disturbance) to 2015 (28 years after
disturbance), 828 388 individual trees and 591 botanical species
pertaining to 223 genus and 64 families were recorded.

While both taxonomic and functional composition remained stable in
undisturbed communities (Fig. \ref{fig:NMDSplans}), they followed marked
and consistent trajectories over post-\break disturbance time. In
disturbed communities, these compositional changes corresponded to
shifts towards species with more acquisitive functional strategies, from
communities with high average WSG to high average SLA and chlorophyll
content (see appendix I). For functional composition, this translated
into cyclic compositional changes with an unachieved recovery of the
initial composition (Fig. \ref{fig:NMDSplans}). The maximum
dissimilarity with the initial state was positively correlated to the
disturbance intensity for both taxonomic and functional composition
(\(\rho_{spearman}^{Taxonomic}=0.87\) and
\(\rho_{spearman}^{Functional}=0.90\) respectively). The maximum value
was reached around 26 years after disturbance for taxonomic composition
and 22 years for functional composition.

\begin{verbatim}
## Loading required package: shape
\end{verbatim}

\begin{verbatim}
## Loading required package: plotrix
\end{verbatim}

{\centering \includegraphics[width=1\linewidth]{Manuscript_files/figure-latex/NMDSplans-1}

\caption{Plot trajectories in terms of flora composition (left panels \textbf{(a)} and \textbf{(c)}) and functional composition (right panels \textbf{(b)} and \textbf{(d)}) in a two-dimensional NMDS space. Lower panels (\textbf{(c)} and \textbf{(d)}) represent the Euclidean distance to initial condition along the 30 sampled years. Shaded areas are the credibility intervals.}\label{fig:NMDSplans}
\end{figure*}

Except for leaf chlorophyll content, which continued to increase for
some T3 and T2 plots 30 years after disturbance, all traits and seed
mass proportions followed unimodal trajectories either stabilizing or
returning towards their initial values.

Maximum height at adult stage (\emph{Hmax}), leaf toughness and wood
specific gravity (\emph{WSG}) first decreased and then slightly
increased but remained significantly lower than their initial value
(Fig. \ref{fig:CWM}). On the other side, bark thickness and specific
leaf area (\emph{SLA}) increased and while bark thickness remained
substantially high after 30 years, \emph{SLA} had almost recovered its
initial value. For all traits, the maximum difference to initial value
was correlated to the disturbance intensity
(\(\rho_{spearman}^{Leaf thickness}=0.76\),
\(\rho_{spearman}^{Chlorophyll content}=0.60\),
\(\rho_{spearman}^{Leaf toughness}=-0.53\),
\(\rho_{spearman}^{SLA}=0.93\), \(\rho_{spearman}^{WSG}=-0.75\),
\(\rho_{spearman}^{Bark thickness}=0.71\),
\(\rho_{spearman}^{Hmax}=-0.40\)). The proportions of the three lightest
seed mass classes increased in all disturbed plots, and decreased after
30 years for the lightest class while it stabilized for the two other
(Supp. Mat. - Fig. S2).

{\centering \includegraphics[width=1\linewidth]{Manuscript_files/figure-latex/CWM-1}

\caption{Trajectories of the communities weighted means over 30 years after disturbance of four leaf traits (Leaf thickness, chlorophyll content, toughness, and specific area), two stem traits (wood specific gravity, and bark thickness) and one life history trait (Specific maximum height at adult stage). }\label{fig:CWM}
\end{figure*}

\subsubsection{Communities richness and
evenness}\label{communities-richness-and-evenness}

For undisturbed plots, taxonomic Richness and Evenness remained stable
over the 30 years monitored. In disturbed communities, after low
disturbance intensity the taxonomic richness increased, reaching a
maximum gain of 14 botanical genera (plot 3 from treatment 2). After
intense disturbance the taxonomic richness followed a more complex
trajectory, decreasing for ten years after disturbance before recovering
to pre-disturbance values. The maximum richness loss or gain after
disturbance was positively correlated to the disturbance intensity
(\(\rho_{spearman}^{Richness}=0.50\)). In all disturbed plots the
taxonomic evenness first increased until a maximum reached after around
20 years. This maximum was positively correlated to the disturbance
intensity (\(\rho_{spearman}^{Evenness}=0.77\)). The evenness then
stabilized except for two T3 plots (plots 8 and 12) for which evenness
kept increasing.

{\centering \includegraphics[width=1\linewidth]{Manuscript_files/figure-latex/DivTaxo-1}

\caption{Trajectories over 30 years of the difference with the 1989 inventory (2 years after disturbance) of community taxonomic \textbf{(a)} richness, \textbf{(b)}, taxonomic evenness, \textbf{(c)} functional richness, and \textbf{(d)} functional evenness. Shaded areas are the credibility intervals }\label{fig:DivTaxo}
\end{figure*}

The plot 7 from treatment 1 displayed constantly outlying functional
richness and evenness and was removed from the graphical representation
for better readability. In undisturbed plots both functional richness
and evenness remained stable along the 30 years. In disturbed plots,
functional richness and evenness trajectories depended on the
disturbance intensity with their maximum positively correlated to \%AGB
loss \(\rho_{spearman}^{Richness}=0.76\) and
\(\rho_{spearman}^{Evenness}=0.60\). Functional richness and evenness
displayed for low disturbance intensity a low but long-lasting increase
up to a maximum reached after 20-25 years, and for high intensity, a
fast but short increase followed after 10 years by a slow decrease
towards the initial values.
